---
title: "Creating residual maps from linear regression"
output: html_document
date: "2023-10-04"
---

1. Remove all objects from R workspace---good programming practice and can help debug your code better.
```{r}   
rm(list = ls())
```

2. Load the "tidyr" package:
```{r}
library(tidyr)
```

3. Read the age adjusted 5-year rates of cancers across the USA. Extract lung cancer rates for California. Order the data by county names.
```{r}
rate_5y <- read.csv("data/age_adjusted.csv")
rate_CA = rate_5y[substr(rate_5y$State_county,1,2) == "CA",]
rate_lung = rate_CA[rate_CA$Site_recode_ICD_O_3_WHO_2008=="Lung and Bronchus",]
rate_lung = rate_lung[order(readr::parse_number(as.character(rate_lung$State_county))),]
```

4. We will need to extract county names from the database. First inspect the "State_county" variable to see the naming pattern.
```{r}
head(rate_lung)
rate_lung$State_county
```

5. Get rid of "CA:" from the names
```{r}
rate_lung$State_county = sapply(strsplit(rate_lung$State_county, ":"), function(x) x[2])
```
6. Get rid of "County" inside the names
```{r}   
rate_lung$State_county = sub("County", "", rate_lung$State_county)
```

7. Get rid of "Registry" inside the names (appears only for "Los Angeles")
```{r}   
rate_lung$State_county = sub("Registry", "", rate_lung$State_county)
```

8. Get rid of "the last 7 characters" from the names (they represent FIPS codes)
```{r}
rate_lung$State_county = substr(rate_lung$State_county, 1, nchar(rate_lung$State_county)-7)
```

9. Get rid of unnecessary white spaces
```{r}   
rate_lung$State_county = trimws(rate_lung$State_county)
```
10. Rename "State_county" as simply "County"
```{r}
names(rate_lung)[names(rate_lung) == "State_county"] <- "County"
```

11. Remove row numbers from the data frame:
```{r}
row.names(rate_lung) <- NULL
```

This completes the creation of the dataframe "rate_lung". 

12. Read the data on smoking rates and inspect its column names.
```{r}
smoking <- read.csv("data/smoking.csv")
head(smoking)
```

13. Rename its variables to "County" and "Smoking_Rate"
```{r}
names(smoking) <- c("County", "Smoking_Rate")
```

14. We will merge the "smoking" data frame with the rate_lung data frame by column "County". It will be important to check for hidden whitespaces in this column for both data frames.
```{r}
smoking$County <- trimws(smoking$County, whitespace="[\t\r\n\\h\\v]")
```

15. Also make sure that Smoking_Rate is numeric
```{r}
smoking$Smoking_Rate <- as.numeric(trimws(sub("%", "", smoking$Smoking_Rate), whitespace="[\t\r\n\\h\\v]"))
```

16. Now we will merge the data into a single data frame 
```{r}
rate_lung <- merge(rate_lung, smoking, by="County")
```

17. Our first model: Run a simple linear regression of "Age_Adjusted_Rate" on "Smoking_Rate"
```{r}
out.lm <- lm(Age_Adjusted_Rate ~ Smoking_Rate, data = rate_lung)
summary(out.lm)
```

18. Obtain the residuals from the linear regression model
```{r}
resid.lm <- residuals(out.lm)
```

19. Plot the residuals on a map: We know how to do this! Begin by loading some libraries. 
```{r}
library(maps)
library(sf)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(classInt)
```

20. Get county map for California from R “maps” library and create an “sf” object.
```{r}
ca.county = map("county","california", fill=TRUE, plot=FALSE)
ca.county.sf <- st_as_sf(ca.county)
```

21. Expand the "sf" object to include the variable you wish to plot. Note that resid.lm is already ordered by county names so it will align.
```{r}
ca.county.sf$resid.lm <- resid.lm
```

22. Create a color palette.
```{r}   
brks_resid.lm = quantile(ca.county.sf$resid.lm, c(0, 0.2, 0.4, 0.6, 0.8, 1))

color.palette = rev(brewer.pal(5,"RdBu"))
class(color.palette)
```

23. Plot the residuals on the map. What do you see?
```{r}   
plot(ca.county.sf["resid.lm"], pal = color.palette, breaks = brks_resid.lm, main = "Lung cancer rates after accounting for age and smoking")
```

