---
title: "Exploratory data analysis for geostatistical data"
author: "Sudipto Banerjee"
output: html_document
date: "2023-10-30"
---

1. Begin by clearing memory.
```{r}
rm(list=ls())
```

2. Load required libraries. "MBA" is an R package for spatial interpolation using multilevel B-spline approximations. The package "fields" is an excellent package containing several functions for exploratory spatial data analysis for point-referenced data. It includes convenient functions for calculating distances between points using planar coordinates as well as great circle distances using spherical coordinates. The package "classInt" helps in creating classes and intervals for plotting category-based data. The package "geoR" offer functions for calculating variograms and carry out spatial model fitting and predictions.
```{r}   
library(MBA)
library(fields)
library(classInt)
library(geoR)
```

3. Read the data set we will explore.
```{r}   
data1=read.csv("../data/dataset1.csv")
```

4. Create a color palette.
```{r}
col.br <- colorRampPalette(c("midnightblue", "cyan", "yellow", "red"))
col.pal <- col.br(5)
```

5. Write a convenient function for plotting an interpolated surface of a column of a data table
```{r}
myplot=function(tab,colname){
  
  surf <- mba.surf(tab[,c("sx","sy",colname)], no.X=100, no.Y=100, h=5, m=2, extend=FALSE)$xyz.est
  image.plot(surf, xaxs = "r", yaxs = "r", xlab="Easting (m)", ylab="Northing (m)", col=col.br(25))
  
}
```

6. Apply the above function to plot surface maps for two variables "y" and "x" in our data.
```{r}   
myplot(data1,"y")
myplot(data1,"x")
```

From the above two plots can you obtain a hint about how y and x are correlated? Let us confirm with a simple linear regression. 

7. Fit a simple linear regression of y on x; obtain the residuals; plot a map of the residuals.
```{r} 
lmobj1=lm(y~x,data=data1)
summary(lmobj1)
data1$res=lmobj1$residuals
myplot(data1,"res")
```

8. Compute an empirical variogram of the dependent variable.
```{r}   
max.dist <- 0.75*max(rdist(data1[,1:2]))
bins <- 20
vario1raw <- variog(coords=data1[,1:2], data=data1$y, uvec=(seq(0, max.dist, length=bins)))
plot(vario1raw,pch=16)
```

9. Compute an empirical variogram of the residuals.
```{r}   
vario1 <- variog(coords=data1[,1:2], data=data1$res, uvec=(seq(0, max.dist, length=bins)))
plot(vario1,pch=16)
```

10. Read a new data set "Western Experimentaal Forest".
```{r}   
WEF.dat=read.csv("../data/WEFsmall.csv")
```

11. Hold out some data for predictions and for Root Mean Square Prediction Errors.
```{r}
set.seed(1) ## Set seed for reproducibility
ind=sample(1:nrow(WEF.dat),100,replace=FALSE) ##sample some rows for holdout
WEF.out=WEF.dat[ind,] ##Holdout or testing data
WEF.in=WEF.dat[-ind,] ##Input or training data
rm("WEF.dat") ##Remove the original data (optional) to free up memory
```

12. Extract some variables: diameter at breast height for the trees and their locations.
```{r}   
DBH <- WEF.in$DBH_cm
coords <- as.matrix(WEF.in[,c("East_m","North_m")])
```

13. Create your color palette. Form intervals based on quantiles and plot the quantiles based on color coding of the locations.
```{r}   
col.br <- colorRampPalette(c("midnightblue", "cyan", "yellow",  "orange", "red"))
col.pal <- col.br(5)
quant <- classIntervals(DBH, n=5, style="quantile")
quant.col <- findColours(quant, col.pal)
plot(coords, col=quant.col, pch=19, cex=1.5, main="", xlab="Easting (m)", ylab="Northing (m)")
legend("topleft", fill=attr(quant.col, "palette"), 
       legend=names(attr(quant.col, "table")), bty="n",cex=1.3)
```

14. Plot an interpolated surface using the MBA package.
```{r}   
surf <- mba.surf(cbind(coords,DBH), no.X=100, no.Y=100, h=5, m=2, extend=FALSE)$xyz.est
image.plot(surf, xaxs = "r", yaxs = "r", xlab="Easting (m)", ylab="Northing (m)", col=col.br(25))
```

15. Plot different species types. Use a new color palette to create color codes based on different species. Each species has its own color palette.
```{r}   
spnum=as.factor(WEF.in$Species)
col.pal2 <- col.br(length(unique(spnum)))
plot(coords, col=col.pal2[spnum], pch=19, cex=1.5, main="", xlab="Easting (m)", ylab="Northing (m)")
legend("topleft", fill=col.pal2, 
       legend=levels(spnum), bty="n")
```

16. Run a simple linear regressiion (ANOVA) on DBH against different species.
```{r}   
lm.DBH <- lm(DBH~Species, data=WEF.in)
summary(lm.DBH)
```

17. Plot a map of the residuals after accounting for spatial variation.
```{r}   
DBH.resid <- resid(lm.DBH)
surf <- mba.surf(cbind(coords,DBH.resid), no.X=100, no.Y=100, h=5, m=2, extend=FALSE)$xyz.est
image.plot(surf, xaxs = "r", yaxs = "r", xlab="Easting (m)", ylab="Northing (m)", col=col.br(25))
```

18, Variogram of raw data and residuals
```{r}
max.dist=0.5*max(rdist(coords))
bins=20
vario.DBH <- variog(coords=coords, data=DBH, uvec=(seq(5, max.dist, length=bins)))
vario.DBH.resid <- variog(coords=coords, data=DBH.resid, uvec=(seq(0, max.dist, length=bins)))
plot(vario.DBH)
plot(vario.DBH.resid)
```

19. Calculate the maximum likelihood estimates of model parameters. Also calculate two model comparison metrics AIC and BIC.
```{r}   
mle <- likfit(coords=coords, data=DBH, trend = trend.spatial(~Species,WEF.in), ini.cov.pars=c(100,100),
              nugget = 300,cov.model="exponential",nospatial=TRUE)
mle
mle$AIC
mle$BIC
```

20. Compare with nonspatial model.
```{r}   
mle$nospatial$AIC
mle$nospatial$BIC
```

21. Now calculate RMSPE from holdout locations.
```{r}   
krig_mlefit=krige.conv(coords=coords, data=DBH,
    locations=WEF.out[,c("East_m","North_m")],krige=krige.control(type.krige="OK",obj.model=mle,
    trend.d=trend.spatial(~Species,WEF.in),trend.l=trend.spatial(~Species,WEF.out)))
```

22. Calculate spatial predictions and RMSPE using holdout locations. 
```{r}   
pred_spatial=krig_mlefit$predict
rmspe_spatial=sqrt(mean((pred_spatial-WEF.out$DBH_cm)^2))

pred_lm=as.vector(as.matrix(trend.spatial(~Species,WEF.out))%*%lm.DBH$coefficients)
rmspe_lm=sqrt(mean((pred_lm-WEF.out$DBH_cm)^2))

rmspe_spatial
rmspe_lm
```

23. Calculate prediction intervals, coverage probabilities (proportion of prediction intervals on holdout sets that include the value at that location) and the width of the confidence interval.
```{r}   
CI_spatial=pred_spatial+1.96*sqrt(krig_mlefit$krige.var)%*%t(c(-1,1))  ## confidence interval ##
CP_spatial=mean(CI_spatial[,1]<WEF.out$DBH_cm & CI_spatial[,2]>WEF.out$DBH_cm) ## coverage probability ##
CIW_spatial=mean(CI_spatial[,2]-CI_spatial[,1]) ## confidence interval width ##

CP_spatial
CIW_spatial
```

24. Compare with predictions from simple linear model
```{r}   
N=nrow(WEF.out)
CI_lm=pred_lm+1.96*summary(lm.DBH)$sigma*cbind(-rep(1,N),rep(1,N))
CP_lm=mean(CI_lm[,1]<WEF.out$DBH_cm & CI_lm[,2]>WEF.out$DBH_cm)
CIW_lm=mean(CI_lm[,2]-CI_lm[,1])

CP_lm
CIW_lm
```

25. Calculate a predicted surface.
```{r}   
WEF.pred=read.csv("../data/WEFpred.csv")

krigsurf_mlefit=krige.conv(coords=coords, data=DBH,
    locations=WEF.pred[,c("East_m","North_m")],krige=krige.control(type.krige="OK",obj.model=mle,
    trend.d=trend.spatial(~Species,WEF.in),trend.l=trend.spatial(~Species,WEF.pred)))

pred=krigsurf_mlefit$predict
predsd=sqrt(krigsurf_mlefit$krige.var)

predsurf <- mba.surf(cbind(WEF.pred[,c("East_m","North_m")],pred), no.X=100, no.Y=100, h=5, m=2, extend=FALSE)$xyz.est
image.plot(predsurf, xaxs = "r", yaxs = "r", xlab="Easting (m)", ylab="Northing (m)", col=col.br(25))
predsdsurf <- mba.surf(cbind(WEF.pred[,c("East_m","North_m")],predsd), no.X=100, no.Y=100, h=5, m=2, extend=FALSE)$xyz.est
image.plot(predsdsurf, xaxs = "r", yaxs = "r", xlab="Easting (m)", ylab="Northing (m)", col=rev(terrain.colors(25)))
```
